Advancements in Moltres for Time-Dependent Multiphysics Molten Salt Reactor Modeling

### 2024-11-23 msre.tex
Scaling test and rod cusping

\begin{figure}[p]
  \centering
  \includegraphics[width=\columnwidth]{msre-quarter-no-rod-power}
  \caption{Normalized channel fission rate distribution of the 2-D \gls{MSRE} quarter-core model.
  This figure depicts the 1/8th-core distribution due to symmetry about the diagonal.}
  \label{fig:1/4-no-rod}
\end{figure}

\begin{figure}[p]
  \centering
  \includegraphics[width=\columnwidth]{msre-quarter-rod-power}
  \caption{Normalized channel fission rate distribution of the 2-D \gls{MSRE} quarter-core model.
  This figure depicts the 1/8th-core distribution due to symmetry about the diagonal.}
  \label{fig:1/4-rod}
\end{figure}

@@

I also calculated the normalized channel fission rate distribution by integrating the fission rate
within each fuel salt channel and normalizing the values such that the integrated fission rate in
each channel averages to 1 as follows
%
\begin{align}
  \text{Normalized fission rate of channel }i &= \frac{\text{Fission rate in channel }i}{
  \text{Total fission rate}}\cdot\frac{\text{Total number of channels}}{
    \text{Total fission rate in channels}} \nonumber \\
                                              &= \frac{\sum^G_{g=1}\int_{V_i}\Sigma_{f,g}
  \phi_g(\vec{r})}{\sum^G_{g=1}\int_V\Sigma_{f,g}\phi_g(\vec{r})}\cdot\frac{N_i}{
  \sum^{N_i}_{i=1}\sum^G_{g=1}\int_{V_i}\Sigma_{f,g}\phi_g(\vec{r})}.
\end{align}
%
The total fission rate includes fission rate contributions from peripheral fuel salt regions.
Figures \ref{fig:1/4-no-rod} and \ref{fig:1/4-rod} show the normalized fission rate distributions
of the quarter-core model for the ``No Rod'' and ``Rod'' cases, respectively. I presented 1/8th
slices of the core in the figures because the distributions are symmetrical about the $y=x$ line
outside of statistical variances for OpenMC-CE. I combined the midpoints between the centroids of
neighboring fuel salt channels to form the diamond lattices shown in the figures.
Each diamond element in the figures corresponds a horizontally or vertically oriented channel
depicted in Figure \ref{fig:1/4-geom}. The lattice also provides a guide for subdividing the
annular fuel regions around the control rod thimbles. The fission rate
distributions omit four channels along the outer edge due to the proximity to peripheral annular
fuel regions and the consequent difficulty in accurately tallying fission rates.

@@

\begin{itemize}
  \item Removed the toroidal flow distributor near the top of the vessel.
  \item Reshaped the upper and lower plena into right cylinders while keeping the salt volume
    constant.
  \item Flattened the conical tops of the graphite lattice blocks.
  \item Regularized the complicated graphite lattice structure at the bottom.
  \item Set a fixed outer vessel thickness of 2.54 cm.
  \item Homogenized the lower plenum by 90.8\% salt- 9.2\% INOR-8 (vessel) by volume.
  \item Removed the reactor access nozzle and the various components within it.
  \item Added a graphite block at the bottom of the control rod thimbles.
\end{itemize}

@@

All neutronics simulations ran on OpenMC under continuous energy mode and on Moltres with the
hybrid $S_N$-diffusion method. The hybrid simulations ran with a maximum stabilization factor of
$c=250$, a void constant $\varsigma=0.5$, and a uncorrected diffusion coefficient cap of $D_g=2.5$.
Refer to Sections \ref{sec:void-treatment} and \ref{sec:diffcoef-cap} for the relevant discussion
on void treatment for near-void regions and the impact of large diffusion coefficient values on
solver performance. As for the standard neutron diffusion method, it either converges very slowly
or does not converge at all when $D_g$ values are uncapped due to excessive diffusive effects in
the air-filled control rod thimbles. Group-wise $D_g$ values, generated using OpenMC, ranged from
100 cm to 850 cm in the air regions as compared to the typical 0.5 cm to 2.0 cm range observed in
non-air regions. Therefore, all neutron diffusion method simulation results in the following
section were measured with the diffusion coefficient capped at $D_g=2.5$.

@@

Beyond demonstrating the hybrid $S_N$-diffusion method, its computational performance is of
interest as a candidate for time-dependent 3-D reactor simulations. All simulations ran on the
Polaris supercomputer system at \gls{ANL}. Each compute node consists of a 2.8-GHz AMD EPYC Milan
7543P 32-core CPU, four NVIDIA A100 GPUs, and 512 GB of RAM. The GPUs were not utilized in this
work. Moltres uses the \gls{MPI} communication protocol for parallelizing simulations. For this
work, simulations ran with 1-to-1 ratio of \gls{MPI} ranks and processor cores (i.e., 32 \gls{MPI}
ranks for each compute node).

Every 3-D hybrid method $k$-eigenvalue
simulation of the \gls{MSRE} consisted of 747,688,760 \glspl{DOF} and 1,690,777,728 \glspl{DOF} for
the neutron diffusion and $S_N$ components, respectively. Each \gls{DOF} corresponds to a scalar
or angular group flux variable value defined at a mesh node. Each simulation ran with the distributed
mesh feature, which partitions and distributes the mesh across however many \gls{MPI} ranks used
in the simulation. All variable and mesh-linked data are also distributed with the mesh. This
greatly reduces memory requirements compared to storing the full mesh for each \gls{MPI} rank.
Each 3-D \gls{MSRE} simulation required at least forty nodes for sufficient memory to run. This
sets a strict minimum requirement for the 3-D hybrid simulations. A significant fraction of
memory use in Moltres comes from storing the residual and Jacobian quantities at each \gls{FEM}
quadrature point. Given that terms in the $S_N$ equations share the same Jacobian formulation
across all ordinate directions, memory use (and compute time) could be optimized by reducing
duplicate Jacobian evaluations.

On forty nodes, the simulations
required approximately 2.2 h to complete for a total of 88 node-hours or 2816 core-hours. The
distributed mesh capability comes with the cost of increased data transfer times between the $S_N$
and diffusion solvers due to increased inter-core data transfers. Transfers accounted for
approximately 0.675 h or 30 \% of the total simulation time. Future research into optimizing mesh
distribution and transfer-related data caching may help to minimize data transfers and speed up
hybrid method simulations. The solution time for the initial neutron diffusion calculation with zero
drift correction also takes up approximately 30 \% of the total simulation time. After subtracting
the minor computational cost of evaluating the drift term even if it is zero, the hybrid method
requires about four times the solution time of the standard neutron diffusion method.

For the hybrid method to be viable for large time-dependent simulations, it must scale reasonably
well on computing clusters. For this work, I investigated strong scaling, which measures how the
solution time decreases when more cores are used while keeping the problem size fixed. This strong
scaling study used 3-D quarter-core models run on 10, 20, 30, and 40 compute nodes.
Figure * shows the results of this scaling study.
The hybrid method implementation scales well as shown by the nearly linear scaling between the
solution time and the number of processors. Therefore, the hybrid method appears promising for the
modeling of larger reactor analyses with little scaling performance penalty.

@@

The 3-D simulations performed well under the strong scaling test which studies how the solution
time scales with the number of processors used to run the simulation. The 3-D simulations showed
close to linear scaling within the scope of the strong scaling test. Therefore, the hybrid method
exhibits good computational scaling performance which is essential for large 3-D reactor analyses.
Excessively large memory requirements were mitigated through the distributed mesh capability in
Moltres. Some performance optimizations may be possible to reduce data transfer times
between the $S_N$ and diffusion solvers which currently take up about 30 \% of the total solution
time as a result of the inter-processor communications.

>>>>

\begin{figure}[p]
  \centering
  \includegraphics[width=\columnwidth]{msre-quarter-no-rod-power}
  \caption{Normalized channel fission rate distribution of the 2-D \gls{MSRE} quarter-core model
    with the rod withdrawn.
  This figure depicts the 1/8th-core distribution due to symmetry about the diagonal.}
  \label{fig:1/4-no-rod}
\end{figure}

\begin{figure}[p]
  \centering
  \includegraphics[width=\columnwidth]{msre-quarter-rod-power}
  \caption{Normalized channel fission rate distribution of the 2-D \gls{MSRE} quarter-core model
    with the rod inserted.
  This figure depicts the 1/8th-core distribution due to symmetry about the diagonal.}
  \label{fig:1/4-rod}
\end{figure}

@@

I also calculated the normalized channel fission rate distribution by integrating the fission rate
within each fuel salt channel and normalizing the values such that the integrated fission rate in
each channel averages to 1 as follows
%
\begin{align}
  \text{Normalized fission rate of channel }i &= \frac{\text{Fission rate in channel }i}{
  \text{Total fission rate}}\cdot\frac{\text{Total number of channels}}{
    \text{Total fission rate in channels}} \nonumber \\
                                              &= \frac{\sum^G_{g=1}\int_{V_i}\Sigma_{f,g}
  \phi_g(\vec{r})}{\sum^G_{g=1}\int_V\Sigma_{f,g}\phi_g(\vec{r})}\cdot\frac{N_i}{
  \sum^{N_i}_{i=1}\sum^G_{g=1}\int_{V_i}\Sigma_{f,g}\phi_g(\vec{r})}.
\end{align}
%
The total fission rate includes fission rate contributions from peripheral fuel salt regions.
Division by total fission rate normalizes the fission rates across the different neutronics
methods.
Figures \ref{fig:1/4-no-rod} and \ref{fig:1/4-rod} show the normalized fission rate distributions
of the quarter-core model for the ``No Rod'' and ``Rod'' cases, respectively. I presented 1/8th
slices of the core in the figures because the distributions are symmetrical about the $y=x$ line
outside of statistical variances for OpenMC-CE. I combined the midpoints between the centroids of
neighboring fuel salt channels to form the diamond lattices shown in the figures.
Each diamond element in the figures corresponds a horizontally or vertically oriented channel
depicted in Figure \ref{fig:1/4-geom}. The lattice also provides a guide for subdividing the
annular fuel regions around the control rod thimbles. The fission rate
distributions omit four channels along the outer edge due to the proximity to peripheral annular
fuel regions and the consequent difficulty in accurately tallying fission rates.

@@

\begin{itemize}
  \item Removed the toroidal flow distributor near the top of the vessel.
  \item Reshaped the upper and lower plena into right cylinders while keeping the salt volume
    constant.
  \item Flattened the conical tops of the graphite lattice blocks.
  \item Regularized the complicated graphite lattice structure at the bottom.
  \item Set a fixed outer vessel thickness of 2.54 cm.
  \item Homogenized the lower plenum by 90.8\% salt-9.2\% INOR-8 (vessel) by volume.
  \item Removed the reactor access nozzle and the various components within it.
  \item Added a graphite block at the bottom of the control rod thimbles.
\end{itemize}

@@

All neutronics simulations ran on OpenMC under continuous energy mode and on Moltres with the
hybrid $S_N$-diffusion method. The hybrid simulations ran with a maximum stabilization factor of
$c=250$, a void constant of $\varsigma=0.5$, and a uncorrected diffusion coefficient cap of $D_g=2.5$.
Refer to Sections \ref{sec:void-treatment} and \ref{sec:diffcoef-cap} for the relevant discussion
on void treatment for near-void regions and the impact of large diffusion coefficient values on
solver performance. As for the standard neutron diffusion method, it either converges very slowly
or does not converge at all when $D_g$ values are uncapped due to excessive diffusive effects in
the air-filled control rod thimbles. Group-wise $D_g$ values, generated using OpenMC, ranged from
100 cm to 850 cm in the air regions as compared to the typical 0.5 cm to 2.0 cm range observed in
non-air regions. Therefore, all neutron diffusion method simulation results in the following
section were measured with the diffusion coefficient capped at $D_g=2.5$.

@@

Beyond demonstrating the hybrid $S_N$-diffusion method, its computational performance is of
interest as a candidate for time-dependent 3-D reactor simulations. All simulations ran on the
Polaris supercomputer system at \gls{ANL}. Each compute node consists of a 2.8-GHz AMD EPYC Milan
7543P 32-core CPU, four NVIDIA A100 GPUs, and 512 GB of RAM. The GPUs were not utilized in this
work. Moltres uses the \gls{MPI} communication protocol for parallelizing simulations. For this
work, simulations ran with 1-to-1 ratio of \gls{MPI} ranks to processor cores (i.e., 32 \gls{MPI}
ranks per compute node).

Every 3-D hybrid method $k$-eigenvalue
simulation of the \gls{MSRE} consisted of 750M \glspl{DOF} and 1.69B \glspl{DOF} for
% 747,688,760 \glspl{DOF} and 1,690,777,728 \glspl{DOF} for
the neutron diffusion and $S_N$ subsolvers, respectively. Each \gls{DOF} corresponds to a scalar
or angular group flux variable value defined at a mesh node. Each simulation ran with the distributed
mesh feature, which partitions and distributes the mesh across however many \gls{MPI} ranks used
in the simulation. All variable and mesh-linked data are also distributed with the mesh. This
greatly reduces memory requirements compared to storing the full mesh for each \gls{MPI} rank.
Each 3-D \gls{MSRE} simulation required at least forty nodes for sufficient memory to run. This
sets a strict minimum requirement for the 3-D hybrid simulations. A significant fraction of
memory use in Moltres comes from storing the residual and Jacobian quantities at each \gls{FEM}
quadrature point. Given that terms in the $S_N$ equations share the same Jacobian formulation
across all ordinate directions, memory use (and compute time) could be optimized by reducing
duplicate Jacobian evaluations.

On forty nodes, each full-core rod worth simulation
required approximately 2.2 h to complete for a total of 88 node-hours or 2816 core-hours. The
distributed mesh capability comes with the cost of increased data transfer times between the $S_N$
and diffusion solvers due to increased inter-processor data transfers. Transfers accounted for
approximately 0.675 h or 30 \% of the total simulation time. Future research into optimizing mesh
distribution and transfer-related data caching may help to minimize data transfers and speed up
hybrid method simulations. The solution time for the initial neutron diffusion calculation with zero
drift correction also takes up approximately 30 \% of the total simulation time. After taking into
consideration the computational cost of evaluating the drift term, the hybrid method
requires about four times the solution time of the standard neutron diffusion method.

For the hybrid method to be viable for large time-dependent simulations, it must scale reasonably
well on computing clusters. For this work, I investigated strong scaling, which measures how the
solution time decreases when more cores are used while keeping the problem size fixed. This strong
scaling study used 3-D quarter-core models run on 10, 20, 40, and 80 compute nodes.
Figure \ref{fig:scaling} shows the results of this scaling study. The total wall time scales at a
sublinear rate of approximately $q=0.7$. The total wall time is broken down into two components:
the solver wall time (i.e., initialization, numerical solver, etc.) and the transfer time (i.e,
data transfer between the $S_N$ and diffusion solvers). The solver wall time scales at a sublinear
rate of approximately $q=0.8$, while the transfer wall time scales at approximately $q=0.6$ from
ten to forty nodes before registering a wall time increase when going from forty nodes to eighty nodes.
It is likely that the increase in inter-processor data transfers due to increased data distribution
outweighed the increase in computational power when running the problem with more compute nodes.
Further investigation is necessary to confirm this hypothesis.

Nevertheless, the total wall time exhibits decent strong scaling performance. For Moltres
simulations in the next chapter, I optimized the data transfer setup which reduced the
proportion of transfer wall time in the total wall time to approximately 18 \%. Therefore, the total
wall times would scale closer to the $q=0.8$ rate observed with solver wall times.
This is a promising indication that the hybrid method would scale well for the modeling of larger
reactor analyses.

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.49\columnwidth}
    \centering
    \includegraphics[width=\columnwidth]{scaling-total}
    \caption{Total wall time}
  \end{subfigure} \\
  \begin{subfigure}[b]{0.49\columnwidth}
    \centering
    \includegraphics[width=\columnwidth]{scaling-solver}
    \caption{Solver wall time}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\columnwidth}
    \centering
    \includegraphics[width=\columnwidth]{scaling-transfer}
    \caption{Transfer wall time}
  \end{subfigure}
  \caption{The total, solver, and transfer wall time of hybrid method simulations of the 3-D
  quarter-core model on 10, 20, 40, and 80 compute nodes (32 processors per node) of the Polaris
  supercomputer. The solver
  time scales almost linearly ($q=0.8$) with the number of processors, while the transfer time
  scales poorly with an increase in wall time at 80 nodes. All axes are in log scale.}
  \label{fig:scaling}
\end{figure}

@@

The 3-D simulations performed well under the strong scaling test which studies how the solution
time scales with the number of processors used to run the simulation. The 3-D simulations showed
close to linear scaling within the scope of the strong scaling test. Therefore, the hybrid method
exhibits good computational scaling performance which is essential for large 3-D reactor analyses.
The simulation setups mitigated excessively large memory requirements through the distributed mesh
capability in Moltres. Some performance optimizations may be possible to reduce data transfer times
between the $S_N$ and diffusion solvers which currently take up about 18 \% of the total solution
time as a result of the inter-processor communications.

### 2024-11-23 transient.tex
Scaling test and rod cusping

>>>>

\subsection{Rod Cusping Effect Correction}

In deterministic codes with spatial discretization, the material interfaces of a partially-inserted
control rod may not always align with mesh elements. This results in a heterogenous
mesh element that is composed of a control rod material at the top and a non-rod material at the
bottom (for control rods that are inserted from above). The rod cusping effect occurs when
homogenizing mixed mesh elements using the volume weighting method to combine dissimilar material
neutron group constants. Rod cusping error typically decreases with increasing axial mesh
resolution because the size of the mixed mesh elements become smaller relative to the entire
problem domain.

Several rod cusping correction methods exist in literature that are variations of the flux-volume
weighting approach to generate more accurate homogenized group constants in the mixed mesh
elements \cite{yamamoto_cell_2004, graham_subplane_2018, schunert_control_2019}. However, due to
a failed flux-volume weighting implementation and time constraints, this work uses an empirical
approach for rod cusping effect correction. The approach involves precomputing the $k_\text{eff}$
with the control rod at various levels of insertion within a mesh element. These simulations use
the simple volume weighting approach to compute a homogenized group constant value as follows:
%
\begin{gather}
  \Sigma_\text{vw} = \frac{V_\text{r} \Sigma_\text{r} + V_\text{nr} \Sigma_\text{nr}}{V_\text{r} + V_\text{nr}} \label{eq:vw}
  \shortintertext{where}
  \begin{align*}
    \Sigma_\text{vw} &= \mbox{ homogenized group constant in mixed mesh element,} \\
    V_\text{r} &= \mbox{ rod volume fraction in mixed mesh element,} \\
    \Sigma_\text{r} &= \mbox{ rod group constant value,} \\
    V_\text{nr} &= \mbox{ non-rod volume fraction in mixed mesh element,} \\
    \Sigma_\text{nr} &= \mbox{ non-rod group constant value.}
  \end{align*}
\end{gather}
%
Note that vertical material and mesh interfaces in the 3-D \gls{MSRE} model always align perfectly
because the control rod moves vertically. Therefore, the mixed mesh elements occur only around the
bottom tip of the control rod. Additionally, the volume fraction is equivalent to the height
fraction, and all mixed mesh elements at the same height share the same volume fraction. Figure
\ref{fig:rod-cusping-1} shows how the fraction of reactivity change changes with the control rod
volume fraction when the control rod is at around mid-reactor and full insertion heights. The
figure depicts the nonlinear relation between the reactivity change and volume fraction that the
``cusping effect'' is named after. Within this small range of rod motion, the relation should be
approximately linear.

\begin{figure}[t]
    \centering
    \begin{subfigure}[t]{.49\textwidth}
        \centering
        \includegraphics[width=\textwidth]{rod-cusping-1}
        \caption{Uncorrected volume weighting}
        \label{fig:rod-cusping-1}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{.49\textwidth}
        \centering
        \includegraphics[width=\textwidth]{rod-cusping-2}
        \caption{Corrected volume weighting}
        \label{fig:rod-cusping-2}
    \end{subfigure}
    \caption{Fraction of reactivity change against control rod volume fraction in mixed mesh
    elements with the rod inserted at mid-reactor height (27.4 to 29.4 inches) and full insertion
    height (0.0 to 1.9 inches).}
    \label{fig:rod-cusping}
\end{figure}

Recognizing that the rod cusping effect approximately follows a $y=x^p$ curve profile, I applied
curve-fitting to obtain the reactivity changes using corrected volume fractions shown in Figure
\ref{fig:rod-cusping-2}. The corrected volume fraction $V_r'$ is calculated as follows
%
\begin{align}
  V_r' &= cV_r^p
  \shortintertext{where}
  p &= 2.5148, \nonumber \\
  c &= 0.43807. \nonumber
  \end{align}
%
The corrected volume fraction produces the desired approximately linear relation between reactivity
change and control rod volume fraction. While this empirical approach worked reasonably well for
this study, a more robust technique will be necessary for future control rod simulations beyond
this work.
